The Impact of Tax Policy Evolution on Economic Development and Income Inequality

### 2024-02-19 paper/paper.qmd
Finish the discussion section

### Pre-process the data for Figure 7

@@

### Pre process for Figure 8

@@

@fig-statutory-tax-rate-vs-gdp-level shows the top statutory tax rate of the personal income tax across 129 countries, utilizing data from 2023 collected by the authors [@article] and particularly distinguishing between high-income countries and low- and middle-income countries. On average, the top statutory tax rate in high-income countries is marginally higher, standing at 35%, compared to 29% in low- and middle-income countries (see @fig-statutory-tax-rate-vs-gdp-level). Despite the slight variation in top statutory tax rates, Figure 8 does not reflect a proportional relationship between GDP per capita and the top statutory tax rate. This underscores the nuanced dynamics of personal income taxation across different income levels and economic contexts.

@@

@@

>>>>

@@

@@

@fig-statutory-tax-rate-vs-gdp-level shows the top statutory tax rate of the personal income tax across 129 countries, utilizing data from 2023 collected by the authors [@article] and particularly distinguishing between high-income countries and low- and middle-income countries. On average, the top statutory tax rate in high-income countries is marginally higher, standing at 35%, compared to 29% in low- and middle-income countries (see @fig-statutory-tax-rate-vs-gdp-level). Despite the slight variation in top statutory tax rates,  @fig-statutory-tax-rate-vs-gdp-level does not reflect a proportional relationship between GDP per capita and the top statutory tax rate. This underscores the nuanced dynamics of personal income taxation across different income levels and economic contexts.

@@

The interplay between the informal sector and tax policy design is pivotal for economic frameworks, particularly in developing economies. @fig-informality-selfemployment and @fig-informality-consumption, showing a marked decline in informality with increased GDP per capita, underscores a critical transition phase for tax policy considerations. This visual evidence aligns with literature indicating that as economies grow, the informal sector's proportionate contribution to the economy tends to diminish, posing both challenges and opportunities for tax policy [@loayza2016informality]. Loayza's work explains how the shrinking of the informal sector as a result of economic development necessitates adaptable tax policies capable of accommodating this transition. Specifically, tax policy effectiveness hinges on recognizing the evolving tax base and ensuring that tax structures are responsive to the changing landscape of formal and informal economic activities.

The implications of these findings are profound. They suggest that policymakers must prioritize the integration of the informal sector into formal economic considerations, tailoring tax policies to be more inclusive and reflective of an economy's developmental stage. This approach not only enhances tax revenue potential but also supports the formalization of informal activities, contributing to broader economic development goals.

### Rethinking Tax Progressivity in Informal Economies

The relationship between economic development, tax revenue, and the progressivity of indirect taxes, such as VAT, undergoes a complex evolution in the presence of significant informal sectors. @fig-taxes-gdp and @fig-taxrevenue-gdp-diff-source serve as empirical backdrops, illustrating the increasing share of tax revenue in tandem with GDP per capita and the progressive nature of tax rates across income deciles, respectively. Contrary to the conventional wisdom that labels indirect taxes as regressive, recent literature presents a nuanced perspective, especially relevant for economies with substantial informal sectors.

Studies by Bird and Gendron @bird2007vat and Alm and Embaye @alm2013taxing challenge the traditional view, providing evidence that VAT and other indirect taxes can exhibit progressive characteristics in contexts where the informal sector plays a dominant role. Bird and Gendron emphasize that the perceived regressivity of VAT is not a given, suggesting that its impact varies significantly across different income groups and is influenced by the structure of the informal sector. Alm and Embaye further this argument by demonstrating that in economies where informal activities are prevalent, the consumption patterns and the tax-burden distribution across different segments of society render indirect taxes less regressive than commonly assumed.

These findings suggest a critical reevaluation of tax policy design in developing economies. They advocate for a nuanced understanding of tax progressivity, one that accounts for the informal sector's impact on consumption and income distribution. By recognizing the progressive potential of indirect taxes in such contexts, policymakers can better tailor tax systems to achieve equity and efficiency goals, even in economies where informal activities are significant.

This literature underscores the importance of context-specific tax policy design, arguing for a more sophisticated analysis that considers the intricate dynamics between formal and informal economies. It points towards the potential for indirect taxes to contribute positively to tax fairness and revenue mobilization efforts in developing countries, challenging entrenched narratives about tax progressivity.

### Dynamics of Indirect Tax Progressivity in Informal Settings

The progressivity of indirect taxes in economies with significant informal sectors is influenced by a myriad of factors, as the second graph of @fig-taxrevenue-gdp-diff-source subtly indicates. This graph, highlighting a shift in the tax structure with economic development, sets the stage for exploring the underlying mechanisms that may contribute to the progressivity of indirect taxes within informal economies. Literature focusing on consumption patterns, tax compliance, and the role of the informal sector provides a comprehensive understanding of these dynamics.

Bahl and Bird @bahl2008should and Ulyssea @ulyssea2018firms offer insights into how consumption patterns within informal sectors can affect the distributional impact of indirect taxes. Bahl and Bird argue that the informal sector's consumption preferences often lean towards goods and services that are either exempt from indirect taxes or are subject to lower rates, which can inadvertently lead to a progressive tax impact. Ulyssea's work complements this view by examining the compliance aspect, suggesting that lower compliance rates in the informal sector for indirect taxes can shift the tax burden towards formal sectors, thereby influencing the overall progressivity of the tax system.

Moreover, the role of the informal sector in the broader economy, as discussed by Jaramillo and Oliva @jaramillo2013taxing, sheds light on the complex interplay between informal activities and tax policy effectiveness. They suggest that the informal sector's adaptability and its interactions with formal economic structures can mediate the impact of indirect taxes on different income groups, contributing to an unintended progressivity in tax outcomes.

These studies collectively underscore the importance of considering the informal sector's unique characteristics when evaluating the progressivity of indirect taxes. They highlight the need for tax policies that are informed by the realities of consumption, compliance, and economic participation in informal settings, suggesting that a nuanced approach to tax design could harness these factors towards more equitable tax outcomes.

@@

The datasets we used in this paper potentially generate ethical concerns and biases, particularly regarding the representation of the informal sector, marginalized populations, such as women, racial and ethnic minorities, low-income households, and individuals working in precarious employment conditions. The ethical concerns emerge from the potential exclusion or underrepresentation of these groups in the economic datasets like the World Bank’s World Development Indicators and labor market data from the International Labour Organization [@article]. This exclusion risks reinforcing systemic inequalities by formulating tax policies based on incomplete pictures of economic realities. Bird and Gendron @bird2007vat highlight the importance of considering the informal sector's impact on tax progressivity, suggesting that overlooking this aspect can lead to biased policy recommendations that fail to address the needs of a significant segment of the population.

Moreover, biases in data collection and analysis processes, as seen in the reliance on formal economic indicators, may not capture the nuanced effects of taxation on different income groups, particularly those within informal economies. The datasets may not fully account for the progressive potential of indirect taxes in environments with significant informality, potentially misrepresenting the tax burden on lower-income households. Grown and Valodia @grown2010taxation emphasize the need for tax policy analysis to include gender-disaggregated data and consider the socio-economic dynamics of informal work to mitigate biases and ensure equitable tax policy formulations.

Addressing these ethical concerns and biases needs a methodological shift towards more inclusive and comprehensive data collection and analysis strategies. Incorporating a broader spectrum of economic activities and demographic details can help capture the diverse impacts of tax policies across different segments of society. This approach aligns with Sen’s @sen1999development advocacy for equity-focused economic analysis, urging researchers to critically examine their methodologies to prevent the perpetuation of existing disparities. By embedding such considerations in tax policy research, scholars can contribute to developing more equitable and effective taxation systems that reflect the complexities of modern economies and foster inclusivity.

### Inclusivity and Representation in Tax Data
The challenge of inclusivity extends to the representation of diverse economic activities and demographic groups within tax data. Policies based on incomplete data risk perpetuating systemic inequities by failing to account for the full spectrum of economic behaviors and conditions (O'Hare, 2018). For instance, tax evasion and avoidance strategies, more accessible to affluent individuals and corporations, are not uniformly represented in tax compliance studies, potentially leading to skewed perceptions of tax fairness and effectiveness.

Ethical Research Practices
Ethical research practices demand a critical examination of these biases and the implementation of methodologies that ensure broad representation and equity in tax policy analysis. Researchers are called upon to adopt inclusive data collection methods, disaggregate data by gender, income, and other relevant demographic factors, and consciously address the systemic biases that may influence their findings (Sen, 1999).

In sum, ethical considerations in tax policy research underscore the importance of conscientious data collection, analysis, and interpretation. By acknowledging and addressing biases, researchers can contribute to the development of tax policies that promote equity and inclusivity across all sectors of society.
